---
title: "bfda_simulations"
author: "Carolina Guidolin"
date: "2025-06-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Running BFDA with final linear mixed model 
The final linear mixed model formulation is:
melatonin suppression = E2 levels + P4 levels + (E2 levels |ID) + (P4 levels | ID) + (1 | ID). 
This means that we are accounting for random intercepts and random slopes. To run this model using the BayesFactor package, we use te notation to include both random slopes and intercept which is specified in [this GitHub issue on the package GitHubpage](https://github.com/richarddmorey/BayesFactor/issues/141#issuecomment-789199943).  

## Importing the informed parameters
These have been calculated in the previous script (informed_data_simulation.Rmd)
```{r}
set.seed(20250602)

library(tidyverse)
library(lmerTest)
library(BayesFactor)
library(furrr)

informed_parameters <- read.csv("informed_parameters.csv")

# Keep only cols of interest
informed_parameters <- informed_parameters %>%
  select(intercept_mean, e2_slope_mean, p4_slope_mean) 
```

## Defining model parameters for bfda simulations
```{r}
# Fixed participant number, dictated by resource limitations
n_ids <- 12 

# Select possible values for intercept mean, based on what we know worked from informed_data_simulation
intercept_mean_list <- list(informed_parameters$intercept_mean)

# Select possible values for the slopes (i.e. the betas of the predictors), based on what we know worked from informed_data_simulation
e2_slope_mean_list <- list(informed_parameters$e2_slope_mean)
p4_slope_mean_list <- list(informed_parameters$p4_slope_mean)

# Fix the standard deviations for the intercept and slopes, based on what we know worked
intercept_sd <- 0.2
e2_slope_sd <- 0.2
p4_slope_sd <- 0.2

# Specify values of E2 and P4 (expressed as standardised values between -1 and 1)
e2_min_value <- -1
e2_max_value <- 1
p4_min_value <- -1
p4_max_value <- 1

# Define number of simulations to run
num_simulations <- 100 

# Define threshold for Bayes Factor 
bf_threshold <- 3
```

# Run parallel simulations using all cores to speed up the simulation
This code will take approximately 5:30 hours on a Windows 11 machine with 32 GB RAM, on 11 cores. 
```{r}
# Use all available cores
plan(multisession, workers = parallel::detectCores() - 1)

# Run BFDA 
new_bfda_simulated_data_parallel_100 <- furrr::future_map_dfr(1:nrow(informed_parameters), function(row) {
  
  # Store BFs for simulations
  bf_list <- numeric(num_simulations)
  
  # Create a for loop to iterate through the possible combinations of parameters (intercept, e2 slopes, and p4 slope)
  for (simulation in seq_len(num_simulations)) {
    
    # Sample intercept and slopes from normal distribution of defined parameters
    intercept <- rnorm(n_ids, mean = informed_parameters$intercept_mean[row], sd = intercept_sd)
    e2_slope  <- rnorm(n_ids, mean = informed_parameters$e2_slope_mean[row], sd = e2_slope_sd)
    p4_slope  <- rnorm(n_ids, mean = informed_parameters$p4_slope_mean[row], sd = p4_slope_sd)
    
    # Simulate data for n_ids individuals
    # Sampling values from a uniform distribution 
    sim_data <- data.frame(
      id = factor(rep(1:n_ids, each = 4)),
      e2_value = runif(4 * n_ids, e2_min_value, e2_max_value),
      p4_value = runif(4 * n_ids, p4_min_value, p4_max_value)
    )
    
    # Create values for y by solving the equation and add noise
    sim_data$y <- intercept[sim_data$id] +
      e2_slope[sim_data$id] * sim_data$e2_value +
      p4_slope[sim_data$id] * sim_data$p4_value +
      rnorm(n_ids * 4, mean = 0, sd = 0.3)
    
    # Keep id separate for later - needed for specifying random slopes
    sim_data$id0 <- sim_data$id 
    
    # Check no created data is NA
    if (anyNA(sim_data)) stop(sprintf("NA found in sim_data at row %d, simulation %d", row, simulation))
    
    # Calculate BF
    # The lmBF function already calculates the ratio between the full model and a model
    # where the intercept is the grand mean
    bf_full <- BayesFactor::lmBF(
      y ~ e2_value + p4_value + id + id0:p4_value + id0:e2_value,
      data = sim_data,
      whichRandom = "id",
      progress = FALSE
    )
    
    # This function calculates the ratio between a model where 
    # the intercept is different for each id compared to a model where the intercept in the grand mean 
    bf_only_intercept <- BayesFactor::lmBF(
      y ~ id,
      data = sim_data,
      whichRandom = "id",
      progress = FALSE
    )
    
    # Take ratio of these two models, meaning the models where intercept
    # is the grand mean cancel each other out
    # So we are effectively taking a ratio between the full model and a model where the intercept is different for id
    bf_ratio <- bf_full / bf_only_intercept
    
    # Extract BF
    bf_list[simulation] <- as.numeric(BayesFactor::extractBF(bf_ratio)$bf)
  }
  
  # Calculate the true positive rate (TPR)
  TPR <- sum(bf_list > bf_threshold) / num_simulations
  
  # Return one row per each iteration
  tibble::tibble(
    intercept_mean = informed_parameters$intercept_mean[row],
    e2_slope_mean  = informed_parameters$e2_slope_mean[row],
    p4_slope_mean  = informed_parameters$p4_slope_mean[row],
    TPR = TPR
  )
},
.options = furrr_options(seed = 20250602) # setting seed so we always get the same results 
)

# Save the data file 
write.csv(new_bfda_simulated_data_parallel_100, "new_bfda_simulated_data_100_parallelised.csv", row.names = FALSE)

```

