# Aim of this script 
This script aims to perform a Bayes Factor Design Analysis (BFDA) for a fixed-n sample. The rationale for this exercise is the need to perform a power analysis for a study on the effect of the menstrual cycle on melatonin suppression, which is currently being designed.

## Why BFDA for power analysis?
The advantage of benefit over frequentist power analysis is that it allows to "plan for compelling evidence". The frequentist hypothesis testing (NSHT) approach does not inform us on the expected strength of evidence of our results, and relies on estimates of effect sizes. This is problematic because extracting effect sizes from the literature is tricky, in the sense that it is rare to find an effect size estimate from a study which had a design very similar to the one we are planning to perform. On the other hand, In fact, the BFDA approach to power analysis allows to answer the question: which evidence strength can I expect for a specific research design? This is quantified with a metric called Bayes Factor (BF). Typically, BFs that are smaller than 1/10 or larger than 10 are counted as strong evidence for the null and alternative hypothesis, respectively.

One implication of BFDA is that a specific study design and model of how the data will be interpreted has to be formulated prior to data collection. In fact, BFDA is performed during the design phase of a study.

## BFDA for the SHINE (sex Steroid Hormones Influence on the Neuroendocrine Effects of light) project
In this study, we are investigating the effect of menstrual cycle phase on melatonin suppression by light at night. Our hypotheses are the following:
H1: In naturally cycling individuals, we hypothesise an effect of menstrual cycle phase on melatonin suppression outcomes.
H0: No effect of menstrual cycle phase on melatonin suppression outcomes.
We believe that the following model described the data:
- Full model: Melatonin AUC relative to dark AUC = β0 +β1 Menstrual phase + β2 (1|participant) + β3 (1|menstrual phase)
- Null model: Melatonin AUC relative to dark AUC = β0 + β1 (1|participant)

The type of BFDA we want to perform in this study is a fixed-n design. This is because we know that we can only have a maximum of n=12 naturally cycling participants for resource and time constraints. Hence, the BFDA can help us to answer this questions:
1. Given a sample size of N=12 and expected population effect size, which Bayes Factors can I expect?
2. What sample size do I need to have to obtain true positive or true negative evidence with a certain probability?

### Fixed-n BFDA workflow
The general workflow we will follow:
1. Generate a full model that you think explains the data, and a related null model
2. Assume a population with certain properties
3. Repeatedly draw random samples from this population
4. Compute the BF for each sample


#### Requirements
- We define a model for the data we are interested in collecting in our experiment. For example y = a + b*x. Let's imagine that y is melatonin suppression response and x is menstrual cycle phase. 
- We define a sample size of n=12
- We assume that a and b have a normal distribution with mean mean_a (5 possible values) and mean_b (5 possible values)
- We assume that the sd of a and the sd of b are fixed, i.e. sd_a = 0.1, sd_b = 0.1
- We also define that x is menstrual cycle phase, so it can only assume 4 values
```{r}
library(tidyverse)
library(BayesFactor)

n_ids <- 12 # participant n
means_a <- seq(0,4,1) # possible means of a 
means_b <- seq(0, 0.4, 0.1) # possible means of b
x_vals <- seq(1,4,1) # menstrual cycle phase
sd_a <- 0.1 # sd of a
sd_b <- 0.1 # sd of b
num_sim <- 1000 # number of simulations
bf_threshold <- 3 # threshold for Bayes Factor
```

### Running the simulations and generating data
We will run 1000 simulations for each combination of mean_a and mean_b. For each simulation, we will generate data for each participant by solving the equation y=a + bx. We will then compute the Bayes Factor for each simulation. We will then calculate the True Positive Rate (TPR) for each combination of mean_a and mean_b. The TPR is defined as the number of BFs that are greater than the threshold divided by the total number of BFs. We will store the results in a dataframe with columns mean_a, mean_b, and TPR. We will then plot this as a heatmap with the x-axis as mean_a, the y-axis as mean_b, and each cell containing the TPR. Fake code as follows:
```{r}
# for each of the 10 mean_a in means_a
#   for each of the 10 mean_b in means_b
#     Run 1000 simulations of the following
#       for each id in n_ids (12)
#         for each of the 4 values of x:
#           #generate data by solving the equation y=a + bx
#           y = rnrom(1, mean_a, sd_a) + rnorm(1, mean_b, sd_b)*x #sampling a and b from their normal distribution
#       get the BF for this one simulation (BayesFactor package compares automatically to identical simulation for the null model, i.e. where the mean_a is 0)
#     get the 1000 BFs
#     Calculate the TPR for that combination of mean_a and mean_b #there will be 100 combinations (10x10), (n bf values > threshold/number of total BFs, in this case 1000) - store this is a df where we have mean_a, mean_b, and TPR as cols 
    
# Plot this as a heatmap with x axis = mean_a, y axis = mean b, and each cell containing the TPR
```

Real code
```{r}
# Store results
results <- data.frame(mean_a = numeric(), mean_b = numeric(), TPR = numeric())

for (mean_a in means_a) {
  for (mean_b in means_b) {
    bf_list <- numeric(num_sim)  # Store BFs for simulations
    print(paste0("Simualtion for mean_a ", mean_a, "and mean_b ", mean_b)) 
    # Run simulations
    for (sim in seq_len(num_sim)) {
      # Generate data for all participants
      a <- rnorm(1, mean_a, sd_a)   # Sample a for each participant
      b <- rnorm(1, mean_b, sd_b)   # Sample b for each participant
      sim_data <- base::expand.grid(id=1:n_ids,x=x_vals)  # Create data frame with all combinations of id and x
      sim_data$y <- a + b*sim_data$x + rnorm(nrow(sim_data),mean=0,sd=0.5)

      # Compute Bayes Factor
      bf <- BayesFactor::lmBF(y ~ x, data = sim_data)
      bf_list[sim] <- as.numeric(BayesFactor::extractBF(bf)$bf)  # Extract BF value 
    }
    
  # Compute True Positive Rate (TPR)
  TPR <- as.numeric(sum(bf_list > bf_threshold) / length(bf_list))
  # Store in results
  results <- rbind(results, data.frame(mean_a = mean_a, mean_b = mean_b, TPR = TPR))
  }
}

# Plot heatmap
ggplot(results,aes(x=mean_a,y=mean_b))+
  geom_tile(aes(fill=TPR), colour = "black")+
  geom_text(aes(label=TPR), colour = "white") +
  labs(title = "True positive rate for different means of a and b parameters",
       x = "Mean a",
       y = "Mean b") +
  coord_fixed(ratio=1) + 
  theme(aspect.ratio = 1)
```
### Suggestions for improvement
```{r}
# If you want to write this code more generally for a given model
# Define a function to act as your model, taking in the parameters
# model_func <- function(a,b,c,d,e,f,g,x) {
#   y <- #model, for example a + b*x
#   return y

# Define the results with all combinations of your parameter means (and sds if you cycle over them)

# results <- expand.grid(mean_a = means_a, mean_b = means_b)
# results$TPR <- NA

#instead of looping over the means (for mean_a in means_a, for mean_b in means_b, etc.) loop over the rows
# for row in results$rows {
#   mean_a <- row$mean_a
#   mean_b <- row$mean_b  ..... etc.

    #instead of writing the function again, just call the model_func you defined above.
    # sim_data$y <- model_func(a,b,x) + noise
# }
```

