---
title: "bfda_visualisations"
author: "Carolina Guidolin"
date: "2025-06-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Visualising TPR for E2 slope, P4 slope and intercept parameter combinations 
First, we need to import the simulated data, which includes the TPR values 
```{r}
simulations_paralleled_100 <- read.csv("new_bfda_simulated_data_100_parallelised.csv")
```

## Heatmaps of pairwise combinations

### Heatmap 1: E2 slope and P4 slope combination, faceted by intercept
```{r}
e2p4_slopes_faceted <- ggplot(simulations_paralleled_100, aes(x = e2_slope_mean, y = p4_slope_mean)) +
  geom_tile(aes(fill = TPR), colour = "black") +
  scale_fill_gradient(low = "white", high = "darkgreen",
                      name = "TPR",
                      breaks = c(0.2, 0.5, 0.8),
                      limits = c(0, 1)
                      ) +
  labs(
    title = "True positive rate (TPR) for E2 and P4 slope means combinations",
    x = "E2 slope mean",
    y = "P4 slope mean"
  ) +
   scale_x_continuous(
    breaks = seq(-0.5, 0.5, by = 0.1),
    limits = c(-0.55, 0.55),
    labels = c("-0.5", "", "-0.3", "", "-0.1", "", "0.1", "", "0.3", "", "0.5"), 
    # labels every second tick to avoid overcrowding
  ) +
  scale_y_continuous(
    breaks = seq(-0.5, 0.5, by = 0.1),
    limits = c(-0.55, 0.55)
  ) +
  coord_fixed(ratio = 1) +
  theme_bw() +
  theme(
    aspect.ratio = 1,
    panel.grid = element_blank(),  # clean up grid lines inside tiles
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12)
  ) + 
  # Changing the labels of each facet so that intercept mean values are rounded to one decimal
facet_wrap( 
  ~ intercept_mean,
  labeller = labeller(
    intercept_mean = function(x) {
      rounded <- round(as.numeric(x), 1)
      # Replace -0.0 with 0.0
      rounded[rounded == -0.0] <- 0.0
      sprintf("%.1f", rounded)
    }
  )
)

# Save plot
ggsave(filename = "heatmap_faceted.png",
       plot = e2p4_slopes_faceted,
       width = 7.5, 
       height = 7.5,
       dpi = 600, 
       bg = "white",
       path= here::here("bfda", "outputs_plots"))
```

### Heatmap 2a: E2 slope and P4 slope, with "collapsed" intercept
Since we observed that the intercept values do not make a big difference, we can now "collapse" the intercept values 
```{r}
e2p4_slopes_single <- ggplot(simulations_paralleled_100, aes(x = e2_slope_mean, y = p4_slope_mean)) +
  geom_tile(aes(fill = TPR), colour = "black") +
  scale_fill_gradient(low = "white", high = "darkgreen",
                      name = "TPR",
                      breaks = c(0.2, 0.5, 0.8),
                      limits = c(0, 1)) +
  labs(
    title = "True positive rate for E2 and P4 slope \nmeans combinations",
    x = "E2 slope mean",
    y = "P4 slope mean"
  ) +
   scale_x_continuous(
    breaks = seq(-0.5, 0.5, by = 0.1),
    labels = c("-0.5", "", "-0.3", "", "-0.1", "0", "0.1", "", "0.3", "", "0.5"),
    limits = c(-0.55, 0.55)
  ) +
  scale_y_continuous(
    breaks = seq(-0.5, 0.5, by = 0.1),
    limits = c(-0.55, 0.55)
  ) +
  coord_fixed(ratio = 1) +
  theme_bw() +
  theme(
    aspect.ratio = 1,
    panel.grid = element_blank(),  
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    title = element_text(size = 16)
  ) 

# Save plot
ggsave(filename = "heatmap_collapsed.png",
       plot = e2p4_slopes_single,
       width = 7, 
       height = 7,
       dpi = 600, 
       bg = "white",
       path= here::here("bfda", "outputs_plots")
)
```

### Heatmap 2b: E2 alope and P4 slope, with "collapsed" intercept and colour coding based on TPR threshold
We colour code blue to be TPR>=0.8 and yellow/cream to be TPR<.08
```{r}
# Create a column for color category
simulations_paralleled_100$TPR_cat <- ifelse(simulations_paralleled_100$TPR >= 0.8, "≥0.8", "<0.8")

e2p4_slopes_single_threshold <- ggplot(simulations_paralleled_100, aes(x = e2_slope_mean, y = p4_slope_mean)) +
  geom_tile(aes(fill = TPR_cat), colour = "black") +
  scale_fill_manual(values = c("≥0.8" = "#66b2d6", "<0.8" = "#ffffbf"), name = "TPR category") +
  labs(
    title = "True positive rate for E2 and P4 slope \nmeans combinations",
    x = "E2 slope mean",
    y = "P4 slope mean"
  ) +
  scale_x_continuous(
    breaks = seq(-0.5, 0.5, by = 0.1),
    labels = c("-0.5", "", "-0.3", "", "-0.1", "0", "0.1", "", "0.3", "", "0.5"),
    limits = c(-0.55, 0.55)
  ) +
  scale_y_continuous(
    breaks = seq(-0.5, 0.5, by = 0.1),
    limits = c(-0.55, 0.55)
  ) +
  coord_fixed(ratio = 1) +
  theme_bw() +
  theme(
    aspect.ratio = 1,
    panel.grid = element_blank(),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    title = element_text(size = 16)
  )

# Save plot
ggsave(filename = "heatmap_collapsed_threshold.png",
       plot = e2p4_slopes_single_threshold,
       width = 5.5, 
       height = 4.5,
       dpi = 600, 
       bg = "white",path= here::here("bfda", "outputs_plots")
       )

```

### Heatmaps 2a and 2b combined
```{r}
library(patchwork)
# Remove titles from individual plots
plot_a <- e2p4_slopes_single + labs(title = NULL)
plot_b <- e2p4_slopes_single_threshold + labs(title = NULL)

# Combine the two plots side by side
combined_plot <- plot_a + plot_b +
  plot_annotation(
    title = "True positive rate (TPR) across E2 and P4 slope means",
    tag_levels = "A",
    theme = theme(
      plot.title = element_text(size = 16, hjust = 0.5)
    )
  ) &
  theme(
    axis.title.y = element_text(size = 14),
    axis.title.x = element_text(size = 14),
    axis.text = element_text(size = 12)
  )

# Save plot
ggsave(filename = "heatmaps_combined.png",
       plot = combined_plot,
       width = 11, 
       height = 4.5,
       dpi = 600, 
       bg = "white",path= here::here("bfda", "outputs_plots")
       )
```


# Description of parameter combinations that lead to TPR values >0.8
Here we provide a summary of the parameter combinations that yielded to TPR values equal to or higher than 0.8.
```{r}

successful_combinations <- simulations_paralleled_100 %>%
  filter(TPR >= 0.8)

successful_p4_slope <- successful_combinations %>%
  count(p4_slope_mean) %>%
  mutate(perc = (n / sum(n)) * 100)

successful_e2_slope <- successful_combinations %>%
  count(e2_slope_mean) %>%
  mutate(perc = (n / sum(n)) * 100)

successful_intercept <- successful_combinations %>%
  count(intercept_mean) %>%
  mutate(perc = (n / sum(n)) * 100)

```

# Description of the FPR, i.e. the TPR when we are simulating under the null model where slope values = 0 
```{r}
fpr_by_intercept <- simulations_paralleled_100 %>%
  filter(e2_slope_mean == 0 & p4_slope_mean == 0) %>%
  summarise(
    FPR = mean(TPR),
    FPR_sd = sd(TPR),
    .groups = "drop"
  )
```