# Aim of this script 
This script aims to perform a Bayes Factor Design Analysis (BFDA) for a fixed-n sample. The rationale for this exercise is the need to perform a power analysis for a study on the effect of the menstrual cycle on melatonin suppression, which is currently being designed.

## Why BFDA for power analysis?
The advantage of benefit over frequentist power analysis is that it allows to "plan for compelling evidence". The frequentist hypothesis testing (NSHT) approach does not inform us on the expected strength of evidence of our results, and relies on estimates of effect sizes. This is problematic because extracting effect sizes from the literature is tricky, in the sense that it is rare to find an effect size estimate from a study which had a design very similar to the one we are planning to perform. On the other hand, In fact, the BFDA approach to power analysis allows to answer the question: which evidence strength can I expect for a specific research design? This is quantified with a metric called Bayes Factor (BF). Typically, BFs that are smaller than 1/10 or larger than 10 are counted as strong evidence for the null and alternative hypothesis, respectively.

One implication of BFDA is that a specific study design and model of how the data will be interpreted has to be formulated prior to data collection. In fact, BFDA is performed during the design phase of a study.

## BFDA for the SHINE (sex Steroid Hormones Influence on the Neuroendocrine Effects of light) project
In this study, we are investigating the effect of menstrual cycle phase on melatonin suppression by light at night. Our hypotheses are the following:
H1: In naturally cycling individuals, we hypothesise an effect of menstrual cycle phase on melatonin suppression outcomes.
H0: No effect of menstrual cycle phase on melatonin suppression outcomes.
We believe that the following model described the data:
- Full model: Melatonin AUC relative to dark AUC = β0 +β1 Menstrual phase + β2 (1|participant) + β3 (1|menstrual phase)
- Null model: Melatonin AUC relative to dark AUC = β0 + β1 (1|participant)

The type of BFDA we want to perform in this study is a fixed-n design. This is because we know that we can only have a maximum of n=12 naturally cycling participants for resource and time constraints. Hence, the BFDA can help us to answer this questions:
1. Given a sample size of N=12 and expected population effect size, which Bayes Factors can I expect?
2. What sample size do I need to have to obtain true positive or true negative evidence with a certain probability?

### Fixed-n BFDA workflow
The general workflow we will follow:
1. Generate a full model that you think explains the data, and a related null model
2. Assume a population with certain properties
3. Repeatedly draw random samples from this population
4. Compute the BF for each sample

#### Requirements
```{r}
# Define a dummy model and its parameters
y = a + bx  # simple linear regression (do not run this line - will throw error)
b_values <- rnorm(50, mean = 0.5, sd = 0.8) # Define a distribution for our "effect size", i.e. for the parameter b 
a <- 3 # for now, we keep a constant because it simplifies calculations
x <- 0.2 # we also keep x constant for now

# Define parameters for our fixed-n design
num_sim <- 500 # number of simulations for each value of a 
bf_threshold <- 3 # bfs which we are accepting for evidence in favour of H1
n <- 12 # n of participants

```

#### Creating a for loop to calculate the power
We want to:
1. Loop over different values of b (effect sizes)
2. For each of them, run 1000 simulations where
- x is randomly sampled from a normal distribution 
- y is calculated using a linear regression formula
- the BF is computed
- All 1000 BFs are stored
3. After the 1000 simulations have been run, we compute power as the proportion of BFs > specific threshold 
4. We store the results in a dataframe 
```{r}
# Store power results (empty for now)
power_results <- numeric(length(b_values))

# Loop over different values of b
for (i in base::seq_along(b_values)) {
  b <- b_values[i]
  bf_list <- numeric(num_sim)

  # Inner loop for 1000 simulations and generation of BFs
  for (j in base::seq_len(num_sim)) {
    # Simulate data
    x <- rnorm(n, mean = 2, sd = 0.8)  # Allow x to vary
    y <- a + b*x + rnorm(n, mean = 0, sd = 1)  # Add some noise 
  
    # Compute Bayes Factor
    data <- data.frame(y = y, x = x)
    bf <- BayesFactor::lmBF(y ~ x, data = data) # should we rather use regressionBF?
   
    # Extract BF value
    bf_extracted <- BayesFactor::extractBF(bf)$bf
    if (is.na(bf_extracted)) {
      print(paste("BF is NA at i =", i, "j =", j))  # Print if BF is NA
    }
    
    bf_list[j] <- bf_extracted
    
  }

  # Compute power as proportion of BFs above threshold
  power_results[i] <- as.numeric(sum(bf_list > bf_threshold) / length(bf_list))
}

# Output results
power <- data.frame(b_values, power_results) %>%
  mutate(power_per = power_results*100)
```

#### Visualising the power based on values in b
```{r}
ggplot2::ggplot() +
  geom_jitter(data = power, aes(x = b_values, y = power_per))
```

### New approach based on Lucien's input

#### Requirements
- We define a model for the data we are interested in collecting in our experiment. For example y = a + b*x. Let's imagine that y is melatonin suppression response and x is menstrual cycle phase. 
- We define a sample size of n=12
- We assume that a and b have a normal distribution with mean mean_a (10 possible values) and mean_b (10 possible values)
- We assume that the sd of a and the sd of b are fixed, i.e. sd_a = 2, sd_b = 5
- We also define that x is menstrual cycle phase, so it can only assume 4 values
```{r}
n_ids <- 12 # participant n
means_a <- seq(0,9,1) # possible means of a 
means_b <- seq(0, 0.9, 0.1) # possible means of b
x_vals <- seq(1,4,1) # menstrual cycle phase
sd_a <- 2 # sd of a
sd_b <- 5 # sd of b
num_sim <- 10
bf_threshold <- 3
```

### Running the simulations and generating data
```{r}
for each of the 10 mean_a in means_a
  for each of the 10 mean_b in means_b
    Run 1000 simulations of the following
      for each id in n_ids (12)
        for each of the 4 values of x:
          #generate data by solving the equation y=a + bx
          y = rnrom(1, mean_a, sd_a) + rnorm(1, mean_b, sd_b)*x #sampling a and b from their normal distribution
      get the BF for this one simulation (BayesFactor package compares automatically to identical simulation for the null model, i.e. where the mean_a is 0)
    get the 1000 BFs
    Calculate the TPR for that combination of mean_a and mean_b #there will be 100 combinations (10x10), (n bf values > threshold/number of total BFs, in this case 1000) - store this is a df where we have mean_a, mean_b, and TPR as cols 
    
Plot this as a heatmap with x axis = mean_a, y axis = mean b, and each cell containing the TPR
```

```{r}

# Store results
results <- data.frame(mean_a = numeric(), mean_b = numeric(), TPR = numeric())

# Loop over values of mean_a and mean_b
for (mean_a in means_a) {
  for (mean_b in means_b) {
    bf_list <- numeric(num_sim)  # Store BFs for simulations
    print(paste0("Simualtion for mean_a ", mean_a, "and mean_b ", mean_b)) 
          
    # Run simulations
    for (sim in seq_len(num_sim)) {
      # Generate data for all participants
      data <- data.frame(
        id = rep(1:n_ids, each = length(x_vals)),  # Repeat each id for 4 x-values
        x = rep(x_vals, times = n_ids),  # Cycle phase values for each participant
        a = rnorm(n_ids, mean_a, sd_a),  # Sample a for each participant
        b = rnorm(n_ids, mean_b, sd_b)   # Sample b for each participant
      )
      data$y <- data$a + data$b * data$x  # Compute y
 
    # Compute Bayes Factor
      bf <- BayesFactor::lmBF(y ~ x, data = data)
      bf_list[sim] <- as.numeric(BayesFactor::extractBF(bf)$bf)  # Extract BF value
      
    # Compute True Positive Rate (TPR)
    TPR <- as.numeric(sum(bf_list > bf_threshold) / length(bf_list))

    # Store in results
    results <- rbind(results, data.frame(mean_a = mean_a, mean_b = mean_b, TPR = TPR))
    }
  }
}
    
```

